---
title: The _"Best"_ Model
---

Most machine learning problems can be solved in various ways. For instance, a *supervised learning* algorithm assigns a label to a new object. To name a few,

-   Logistic Regression
-   *Stochastic Gradient Descent (SGD)*
-   K-nearest Neighbors
-   Naïve Bayes
-   *Support Vector Machine (SVC)*
-   Decision Tree

Training a machine learning model may require one or more parameters be presumed beforehand, referred to as `hyperparameters`. Finding the best model in a specific use case has led to research and development of model selection, tuning and optimization, and aggregation techniques.

There are at least two problems to be solved.

::: {.callout-important appearance="simple"}

### The Two Problems

1. Firstly, for a given dataset, which algorithm(s) would perform best for a specific use case in terms of chosen metrics? 

2. Each algorithm has its peculiar hyperparameters. After determining the best algorithm candidates, how do we search for the best set of hyperparameters?  

:::

For instance,

::: panel-tabset

### K-nearest Neighbors

The number of neighbors $K$



### Linear Model

The number of independent features $n$ in a linear model

$$
y = b + a_1x_1 + \dots + a_nx_n
$$


### Decision Tree

-   The maximum depth, i.e., height of tree
-   The minimum number of samples required in a leaf node


```{r}
#| echo: false
#| column: screen-inset-shaded
#| layout-nrow: 1
library(ISLR)
library(rpart)
library(rpart.plot)
#build the initial decision tree
tree <- rpart(Salary ~ Years + HmRun, data=Hitters, control=rpart.control(cp=.0001))

rpart.plot(tree)

#identify best cp value to use
best <- tree$cptable[which.min(tree$cptable[,"xerror"]),"CP"]

#produce a pruned tree based on the best cp value
pruned_tree <- prune(tree, cp=best)
#plot the pruned tree
prp(pruned_tree)
```

### Random Forest

-   The number of trees in a random forest



### Gradient Descent

-   The learning rate $\alpha \in [0.0, 1.0]$ to control the speed of learning

$$
\alpha \times \nabla
$$

### Neural Network

- The number of layers 
- The number of neurons per layer 

:::


This lesson discusses the techniques relevant to model section and tuning in a single-model scenario. A model selection technique trains multiple models and compares their performance for a given dataset. A model optimization technique determines the best configuration for a specific model.

Besides replying on a single algorithm, it is common practice to run two or more algorithms and consolidate the results in the final output, referred to as **ensemble modeling**.


## ExampleSet

We randomly generate a numerical dataset of $N$ samples and train a binary classifier from the samples. The example data contains two independent features $x1$ and $x2$, and one class label $y$. We use the same example set for the model selection and tuning parts in this lesson.

::: callout-tip

A one-third hold-out ratio is used to separate data into the training and test sets. The subsequent process for model selection is performed on the training set, `X_train` and `y_train` only.

:::

```{python}
#| results: hold
#| eval: true

# 1. Data Generation 
import numpy as np
# configuration
np.random.seed(99) 
labels = [0, 1]
multiplier = 20
N = 5 * multiplier
N_folds = 5 # CV
holdout = 1/3 # ratio 
# A dataset of 2 features x1, x2 & 1 label y
y = np.random.choice(labels, size=N, replace=True)
x1 = np.array([0.1, 0.2, 0.3, 0.4, 0.5] * multiplier)
x2 = np.random.sample(N)
X = np.stack((x1, x2), axis=-1)

# 2. Holdout 70% for training, 30% for test
from sklearn.model_selection import StratifiedKFold, train_test_split, cross_validate
X_train, X_test, y_train, y_test = train_test_split(X, y, test_size=holdout, shuffle=True, random_state=0)

# class distribution
print(f'''
Head:
{X[0:5]}
CLASS DISTRIBUTION after Stratified Split
y:
Percentage of label 0: {round(sum(y==0)/len(y),5)}
Percentage of label 1: {round(sum(y==1)/len(y),5)}
y_train:
Percentage of label 0: {round(sum(y_train==0)/len(y_train),5)}
Percentage of label 1: {round(sum(y_train==1)/len(y_train),5)}  
y_test:
Percentage of label 0: {round(sum(y_test==0)/len(y_test),5)}
Percentage of label 1: {round(sum(y_test==1)/len(y_test),5)}  
''')
```




## Model Selection

Without model comparison, it is hard to anticipate the best performed model type for a given dataset. Empirical experience can firstly be used to initially filter the candidates. The selection process can subsequently be executed to compare the candidates.


### Determine Candidates


::: panel-tabset

#### Use the Estimator Diagram

The following is a useful diagram for choosing the right type of estimator based on the three conditions: `data type`, `data size` and `learning type`.

```{r echo=F}
knitr::include_url("https://scikit-learn.org/stable/tutorial/machine_learning_map/index.html")
```

#### To Determine Model Candidates

1.  KNeighbors Classifier
2. Logistic Regression
3.  SGD Classifier (_Stochastic Gradient Descent_)
4.  SVC (_Support vector machine_)


:::

### Compare Candidates


::: panel-tabset
#### Problem Case

Use Scikit-Learn to compare the classification algorithms determined from the last step (*All models are trained with default parameters*.):

1.  `KNeighborsClassifier`
2. `LogisticRegressioin`
2. `SGDClassifier`
3.  `SVC` (Support vector machine)


#### Code: Scikit-Learn APIs

```{python}
#| results: hold
#| eval: true

# continue
# 3. Determine Model Candidates
from sklearn.neighbors import KNeighborsClassifier
from sklearn.linear_model import LogisticRegression, SGDClassifier
from sklearn.svm import SVC
from sklearn.naive_bayes import GaussianNB

# 4. group estimator names and objects in a dictionary
estimators = {
  'KNeighborsClassifier': KNeighborsClassifier(algorithm='auto',n_neighbors=5), 
  'LogisticRegression': LogisticRegression(random_state=0),
  'SGDClassifier': SGDClassifier(alpha=0.00001, max_iter=1000, early_stopping=False, tol=0.001),
  'SVC': SVC(gamma='scale')
  }

for estimator_name, estimator_object in estimators.items():
  # kfold = KFold(n_splits=2, random_state=11, shuffle=True)
  scores = cross_validate(estimator_object, X_train, y_train, cv=N_folds, return_train_score=True, return_estimator=True)
  print(f'{estimator_name:>20}: ' + 
           f'mean accuracy={scores["test_score"].mean():.2%}; ' +
           f'standard deviation={scores["test_score"].std():.2%}')
```

:::


### Question

How would you modify the code given above so that it would also test a Naive Bayes classifier model, `'GaussianNB': GaussianNB()`?


## Model Tuning

Once we have identified a shortlist of models in terms of specific metrics, we should fine-run them to find hyperparameter values that produce the best possible results.

::: callout-tip
Model tuning runs K-fold cross validation over the training data.
:::

### Grid Search

The Grid search technique is a brute-force method which explores every combination globally in one-/multi-dimensional parameter space, do model evaluation and search for the best parameter values.

### 1D Grid

::: panel-tabset

#### Search for best K for K-nearest Classifier

Get the samples

Define search grid and its granularity by K_min, K_max, and step_size

For each value $ \in [K_min, K_max, step_size]$ 
  
  Do K-fold cross validation
  
  record k and its metric

Find the k value having the best metric


#### Grid Search from Scratch

Not required

`numpy.arange`

`numpy.meshgrid`


#### Scikit-Learn model_selection.GridSearch API

```{python}
#| results: hold
#| 
# Grid search w/ CV
import numpy as np
from time import time
import scipy.stats as stats

from sklearn.neighbors import KNeighborsClassifier
from sklearn.model_selection import StratifiedKFold, GridSearchCV

# Utility function to report best scores
def report(results, n_top=3):
  for i in range(1, n_top + 1):
    candidates = np.flatnonzero(results["rank_test_score"] == i)
    for candidate in candidates:
      print(f'''
      Model with rank = {i}
      Mean validation score: {results["mean_test_score"][candidate]:.3f} 
      std: {results["std_test_score"][candidate]:.3f}
      Parameters: {results["params"][candidate]}
      ''')

knn = KNeighborsClassifier()
# Define grid
param_grid = {
    "n_neighbors": np.arange(1,21)
}
N_folds = 5
# run grid search
grid_search = GridSearchCV(knn, param_grid=param_grid, cv=StratifiedKFold(n_splits=N_folds, shuffle=False))
start = time()
grid_search.fit(X_train, y_train)
#
print(f'''
GridSearch with {N_folds}-Fold CV took {time()-start:.2f} seconds for {len(grid_search.cv_results_["params"])} candidate parameter settings.
Best_estimator: {grid_search.best_estimator_}
''')

report(grid_search.cv_results_)

```

#### Plot `K` vs. `mean_test_score`

```{python}
#| results: hold
import matplotlib.pyplot as plt
import numpy as np
np.random.seed(99)

ks = grid_search.cv_results_["param_n_neighbors"]
scores = grid_search.cv_results_["mean_test_score"]
print(ks, scores)
colors = np.random.rand(len(ks))
area = scores * 20
plt.scatter(ks, scores, s=area, c=colors, alpha=0.5)
#plt.plot(np.append(scores[0],scores), linestyle="dotted")
plt.title('The change of mean_test_score with $K$ values')
plt.ylabel('mean_test_score')
plt.show()
```


#### Object names in `cv_results_`

```{python}
for obj in grid_search.cv_results_:
  print(obj)

```




:::


### 2D-Grid Search

Tuning two parameters simultaneously requires a 2-dimensional grid. 

::: panel-tabset

#### Tuning learning_rate and stopping condition for SGD Classifier

To tune a SGD classifier, define a 2-dimensional grid for the learning rate $alpha$ and stopping condition $tol$.


```{python}
#| eval: false
n_alpha, n_tol = 10, 5
param_grid = {
    "alpha": np.linspace(0.6, 1, n_alpha),
    "tol": np.linspace(0.001, 0.01, n_tol)
}
print(param_grid)
```

#### Scikit-Learn Code

```{python}
#| results: hold
#| 
# Grid search w/ CV
import numpy as np
from time import time
import scipy.stats as stats

from sklearn.linear_model import SGDClassifier
from sklearn.model_selection import GridSearchCV

# default: alpha=0.00001, max_iter=1000, early_stopping=False, tol=0.001
sgd = SGDClassifier()
# Define grid
n_alpha, n_tol = 20, 10
param_grid = {
    "alpha": np.linspace(0.6, 1, n_alpha),
    "tol": np.linspace(0.01, 0.1, n_tol)
}
N_folds = 5
# run grid search
grid_search = GridSearchCV(sgd, param_grid=param_grid, cv=StratifiedKFold(n_splits=N_folds, shuffle=False))

start = time()
grid_search.fit(X_train, y_train)

```

#### Get Best Estimator

```{python}
#| results: hold


# Utility function to report best scores
def report(results, n_top=3):
  for i in range(1, n_top + 1):
    candidates = np.flatnonzero(results["rank_test_score"] == i)
    for candidate in candidates:
      print(f'''
      Model with rank = {i}
      Mean validation score: {results["mean_test_score"][candidate]:.3f} 
      std: {results["std_test_score"][candidate]:.3f}
      Parameters: {results["params"][candidate]}
      ''')
      
print(f'''
GridSearch with {N_folds}-Fold CV took {time()-start:.2f} seconds for {len(grid_search.cv_results_["params"])} candidate parameter settings.
Best_estimator: {grid_search.best_estimator_}
''')

report(grid_search.cv_results_, n_top=1)
```


#### Object Names in `cv_results_`

```{python}
for obj in grid_search.cv_results_:
  print(obj)

```

#### Contour Plot `(alpha,tol)` vs. `mean_test_score`

```{python}
#| eval: true
#| results: hold

# Display 2d data in a 2d plot
# Display 2d data in a 2d plot
import matplotlib.pyplot as plt
import numpy as np
np.random.seed(99)

alpha = np.linspace(0.6, 1, n_alpha)
tol = np.linspace(0.1, 0.01, n_tol)
scores = grid_search.cv_results_["mean_test_score"]
print(scores[0:5])

A, T = np.meshgrid(alpha, tol)
print(A.shape, T.shape)
# 
Z = np.reshape(np.flip(scores), (n_tol, n_alpha))
plt.clf()
plt.contourf(A, T, Z, 20, cmap='RdGy') #RdGy
plt.colorbar()
plt.title('How does mean_test_score vary with $(alpha, tol)$?')
plt.show()

```

:::




### Question

How would you modify the code given above so that it would fine-tune the following two hyperparameters for SGD classifier?

- alpha \in [0.5, 1] 
- max_iter \in []
- 


